A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-19 podc_paper/algorithm.tex
Working on conclusion section

We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each node of the tree stores an array of \emph{blocks} and each \emph{block} represents a 
sequence of enqueues and a sequence of dequeues.
To perform an operation on the queue, a process $P$ creates a block containing that single 
operation and appends it to the array in $P$'s leaf.
Then, $P$ attempts to propagate the operation along the path from that leaf to the root of the tree.
At each node $v$ along this path, $P$ combines information from the blocks in $v$'s children into a single
block and attempts to append this block to $v$'s array using a CAS instruction.
This cooperative approach is sufficient to ensure that if $P$ fails to CAS a new block into $v$'s array twice,
then its operation has been propagated to $v$ by some other process.
We define a total order on all operations that have been propagated to the root, which is used to
linearize the operations.
In this section, we assume for simplicity that each node has an infinite blocks array.
We describe in Section \ref{??} how to replace the infinite array by a representation that uses bounded space.

@@

>>>>

We use a \emph{tournament tree} to agree on a total ordering of the operations performed on the queue.
The tree is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process. 
Each node of the tree stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
To perform an operation on the queue, a process $P$ creates a block containing that single 
operation and appends it to the array in $P$'s leaf.
Then, $P$ attempts to propagate the operation along the path from that leaf to the root of the tree.
At each node $v$ along this path, $P$ combines information from the blocks in $v$'s children into a single
block and attempts to append this block to $v$'s array using a CAS instruction.
This cooperative approach is sufficient to ensure that if $P$ fails to CAS a new block into $v$'s array twice,
then its operation has been propagated to $v$ by some other process.
We define a total order on all operations that have been propagated to the root, which is used to
linearize the operations.
In this section, we assume for simplicity that each node has an infinite blocks array.
We describe in Section \ref{??} how to replace the infinite array by a representation that uses bounded space.

@@

\here{At what point would a figure illustrating of the tournament tree be useful?}

### 2022-12-19 podc_paper/podc-submission.tex
started working on description of the algorithm
Working on conclusion section

\newcommand{\here}[1]{\bf{[[#1]]}}
\newcommand{\ceil}[1]{\lceil #1 \rceil}

@@

\pagebreak
\section{Future Directions}
We designed a tree to achieve agreement on a linearization of operations invoked by $p$ processes in an asynchronous model, which we will call a \it{block tree}. We implemented two queries to compute information about the ordering agreed in the block tree. Then we used the tree to implement a queue where the number of steps per operation is poly-logarithmic with respect to the size of the queue and the number of processes. Block trees can be used as a mechanism to achieve agreement among processes to construct more poly-logarithmic wait-free linearizable objects. In the next paragraphs, we talk about possible improvements on block trees and the data structures that we can implement with block trees.

\paragraph{Reducing Space Usage}
The \nf{blocks} arrays defined in our algorithm are unbounded. To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}. We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}). When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not. If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}. Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array. When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps. The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}. Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}. Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps. The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$. The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
\begin{figure}[hbt]  
  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
  \caption{Array segments.}
  \label{fig::doublingArray}
\end{figure}

@@

\paragraph{Poly-logarithmic Wait-free Sequences}
Consider a data structure storing a sequence that supports three operations \nf{append(e)}, \nf{get(i)} and \nf{index(e)}. An \nf{append(e)} adds \nf{e} to the end of the sequence, a \nf{get(i)} gets the \nf{i}th element in the sequence and an \nf{index(e)} computes the position of element \nf{e} in the sequence. We can modify our queue to design such a data structure. An \nf{append(e)} is implemented like \nf{Enqueue(e)}, \nf{get(i)} is done by calling \nf{DoublingSearch} but with a \nf{BinarySearch} on the entire \nf{root.blocks} array and \nf{index(e)} is done similarly to \nf{IndexDequeue} (except operating on enqueues instead of dequeues).
 We achieve this with poly-logarithmic steps for each operation with respect to the number of \nf{append}s done.

\paragraph{Other Poly-log Wait-free Data structures}
There are two reasons the block tree worked well to implement a queue. Firstly, to respond to a \nf{Dequeue} we do not need to look at the entire history of operations: if a \nf{Dequeue} does not return \nf{null}, we can compute the index of the \nf{Enqueue} that is its response in $O(\log n)$ time if we keep the number of enqueues and the size. Secondly, the operations we need to search to respond to the \nf{Dequeue} are not very far from it in the sequence of operations: the distance is at most linear in the size of the queue. Creating a wait-free poly-logarithmic implementation of other objects whose operations satisfy these two conditions may be possible.

\paragraph{Attiya--Fouren Lower Bound}
As discussed in Section 2.4 the Attiya--Fouren lower bound says that a concurrent implementation of queues using reads, writes and conditional operations like \nf{CAS} has $\Omega(c)$ amortized complexity \cite{DBLP:conf/opodis/AttiyaF17} when $c$ the number of concurrent processes, is $O(\log\log p)$. Our amortized worst-case step complexity is $\Theta(\log^2p+\log q)$. It is an open problem to reduce the gap between our algorithm and the Attiya--Fouren lower $\Omega(\log\log p)$ bound.

>>>>

\newcommand{\here}[1]{\bf{[[#1]]}}
\newcommand{\ceil}[1]{\lceil #1 \rceil}
\newcommand{\nil}{nil}
\newcommand{\opa}[2]{\nf{#1(#2)}}
\newcommand{\op}[1]{\nf{#1}}
\newcommand{\var}[1]{\nf{#1}}

@@

\section{Reducing Space}
\label{reducing}
\paragraph{Reducing Space Usage}
The \nf{blocks} arrays defined in our algorithm are unbounded. To use $O(n)$ space in each node where $n$ is the total number of operations, instead of unbounded arrays, we could use the memory model of the wait-free vector introduced by Feldman, Valera-Leon and Damian~\cite{7073592}. We can create an array called \nf{arr} of pointers to array segments (see Figure \ref{fig::doublingArray}). When a process wishes to write into location \nf{head} it checks whether \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array or not. If not, it creates a shared array of size $2^{\lfloor\log \nf{head} \rfloor}$ and tries to \nf{CAS} a pointer to the created array into \nf{arr[$\lfloor\log $ head$\rfloor$]}. Whether the \nf{CAS} is successful or not, \nf{arr[$\lfloor\log $ head$\rfloor$]} points to an array. When a process wishes to access the $i$th element it looks up \nf{arr[$\lfloor\log i\rfloor$][$i-2^{\lfloor\log i\rfloor}$]}, which takes $O(1)$ steps. The CAS Retry Problem does not happen here because if $n$ elements are appended to the array, then only $O(p\times\log n)$ \nf{CAS} steps have happened on the array \nf{arr}. Furthermore, at most $p$ arrays with size $2^{\lfloor\log i\rfloor}$ are allocated by processes while processes try to do the \nf{CAS} on \nf{arr[$i$]}. Jayanti and Shun \cite{DBLP:conf/wdag/JayantiS21} present a way to initialize wait-free arrays in constant steps. The time taken to allocate arrays in an execution containing $n$ operations is $O(\frac{p\log n}{n})$  per operation, which is negligible if $n>>p$. The vector implementation also has a mechanism for doubling \nf{arr} when necessary, but this happens very rarely since increasing \nf{arr} from $s$ to $2s$ increases the capacity of the vector from $2^s$ to $2^{2s}$.
\begin{figure}[hbt]  
  \center\includegraphics[width=6.5in]{pics/doublingArray.png}
  \caption{Array segments.}
  \label{fig::doublingArray}
\end{figure}

@@

\input{conclusion}
